Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2021-02-19 summary.Rmd
very rough proofs

Theorem 2 suggests that SSC is appropriate for community detection for the 
PABM. More precisely, theorem 2 says that each community consists of a
$K$-dimensional subspace, and together the subspaces lie in $\mathbb{R}^{K^2}$. 
The natural approach then is to perform SSC on the ASE of $P$ or $A$.
\citet{noroozi2019estimation} instead applied SSC to $P$ and $A$ themselves.

@@

Theorems 2, 3, and 4 also provide a very natural path toward using SSC for 
community detection for the PABM. We established in theorem 2 that an ASE of 
the edge probability matrix $P$ can be constructed such that the communities 
lie on mutually orthogonal subspaces, and this property can be recovered from
the eigenvectors of $P$. Then theorems 3 and 4 show that this property holds 
for the adjacency matrix $A$ drawn from $P$ as $n \to \infty$.

@@

For any edge probability matrix $P$ for the PABM such that the rows and 
columns are organized by community, the $kl$^th^ block is an outer product 
of two vectors, i.e., $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. 
Therefore, given $P^{(kl)}$, $\lambda^{(kl)}$ and $\lambda^{(lk)}$ are 
solvable exactly (up to multiplication by $-1$) using singular value 
decomposition. More specifically, let $P^{(kl)} = \sigma^2 u v^\top$ be the 
singular value decomposition of $P^{(kl)}$. $u \in \mathbb{R}^{n_k}$ and 
$v \in \mathbb{R}^{n_l}$ are vectors and 
$\sigma^2 > 0$ is a scalar. Then $\lambda^{(kl)} = \pm \sigma u$ and 
$\lambda^{(lk)} = \pm \sigma v$. 
Given the adjacency matrix $A$ instead of edge probability matrix $P$, we can 
simply use plug-in estimators (algorithm 3), which converge to the true 
parameters (theorem 6).

@@

**Lemma 1.** 
Let $P = V D V^\top$ be the spectral decomposition of the edge probability 
matrix for a PABM. Then $V V^\top = X (X^\top X)^{-1} X^\top$ where $X$ is 
defined as in equation (7).

*Proof of Lemma 1*. 
By Theorem 2, $P = X U I_{p, q} U^\top X^\top$, where $X$ is defined as in 
equation (7) and $p$ and $q$ are defined as in equations (10) and (11).
Alternatively, the spectral decomposition can be written as 
$P = V D V^\top = V |D|^{1/2} I_{p, q} |D|^{1/2} V^\top$ for the same $(p, q)$ 
and $|\cdot|^{1/2}$ is applied entry-wise. Thus for some 
$Q \in \mathbb{O}(p, q)$,

@@

**Proof of Theorem 3**.
By Lemma 1, $V V^\top = X (X^\top X)^{-1} X^\top$ where $X$ is defined as 
in Theorem 2. Since $X$ is block diagonal with each block corresponding 
to one community, $X (X^\top X)^{-1} X^\top$ is also a block diagonal matrix 
with each block corresponding to a community and zeros elsewhere. Therefore, 
if vertices $i$ and $j$ belong to different communities, then the $ij$^th^ 
element of $X (X^\top X)^{-1} X^\top = V V^\top = B$ is 0.

**Proof of Theorem 5**.

>>>>

Theorem 2 suggests that SSC is appropriate for community detection for the 
PABM. More precisely, Theorem 2 says that each community consists of a
$K$-dimensional subspace, and together the subspaces lie in $\mathbb{R}^{K^2}$. 
The natural approach then is to perform SSC on the ASE of $P$ or $A$.
\citet{noroozi2019estimation} instead applied SSC to $P$ and $A$ themselves.

@@

Theorems 2, 3, and 4 also provide a very natural path toward using SSC for 
community detection for the PABM. We established in Theorem 2 that an ASE of 
the edge probability matrix $P$ can be constructed such that the communities 
lie on mutually orthogonal subspaces, and this property can be recovered from
the eigenvectors of $P$. Then Theorems 3 and 4 show that this property holds 
for the adjacency matrix $A$ drawn from $P$ as $n \to \infty$.

@@

For any edge probability matrix $P$ for the PABM such that the rows and 
columns are organized by community, the $kl$^th^ block is an outer product 
of two vectors, i.e., $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. 
Therefore, given $P^{(kl)}$, $\lambda^{(kl)}$ and $\lambda^{(lk)}$ are 
solvable exactly (up to multiplication by $-1$) using singular value 
decomposition. More specifically, let $P^{(kl)} = \sigma^2 u v^\top$ be the 
singular value decomposition of $P^{(kl)}$. $u \in \mathbb{R}^{n_k}$ and 
$v \in \mathbb{R}^{n_l}$ are vectors and 
$\sigma^2 > 0$ is a scalar. Then $\lambda^{(kl)} = \pm \sigma u$ and 
$\lambda^{(lk)} = \pm \sigma v$. 
Given the adjacency matrix $A$ instead of edge probability matrix $P$, we can 
simply use plug-in estimators (algorithm 3), which converge to the true 
parameters (Theorem 6).

@@

**Lemma 3.1.** 
Let $P = V D V^\top$ be the spectral decomposition of the edge probability 
matrix for a PABM. Then $V V^\top = X (X^\top X)^{-1} X^\top$ where $X$ is 
defined as in equation (7).

*Proof of Lemma 3.1*. 
By Theorem 2, $P = X U I_{p, q} U^\top X^\top$, where $X$ is defined as in 
equation (7) and $p$ and $q$ are defined as in equations (10) and (11).
Alternatively, the spectral decomposition can be written as 
$P = V D V^\top = V |D|^{1/2} I_{p, q} |D|^{1/2} V^\top$ for the same $(p, q)$ 
and $|\cdot|^{1/2}$ is applied entry-wise. Thus for some 
$Q \in \mathbb{O}(p, q)$,

@@

**Proof of Theorem 3**.
By Lemma 3.1, $V V^\top = X (X^\top X)^{-1} X^\top$ where $X$ is defined as 
in Theorem 2. Since $X$ is block diagonal with each block corresponding 
to one community, $X (X^\top X)^{-1} X^\top$ is also a block diagonal matrix 
with each block corresponding to a community and zeros elsewhere. Therefore, 
if vertices $i$ and $j$ belong to different communities, then the $ij$^th^ 
element of $X (X^\top X)^{-1} X^\top = V V^\top = B$ is 0.

**Lemma 5.1**. 
Let $P_n = V_n \Lambda_n V_n^\top$ be defined as in Lemma 5.1. Let 
$A_n \sim \Bernoulli(P_n)$ be the adjacency matrix drawn from $P_n$, and let 
$A_n = \hat{V}_n \hat{\Lambda}_n \hat{V}_n^\top$ be its approximate spectral
decomposition using the $K (K+1) / 2$ most positive and $K (K-1) / 2$ most 
negative eigenvalues and their corresponding eigenvectors.
Without loss of generality, order the rows of $V_n$ and $\hat{V}_n$ by 
community, and let $V_n^{(k)}$ and $\hat{V}_n^{(k)}$ be the $n_k \times K^2$ 
matrix of the rows of the $k$^th^ community in $V_n$ and $\hat{V}_n$ 
respectively. Let $r_n^{(k)} = r(V_n^{(k)})$ and 
$\hat{r}_n^{(k)} = r(\hat{V}_n^{(k)})$ be the inradius of points in $V_n$ and
$\hat{V}_n$ as defined by \citet{soltanolkotabi2012} and 
\citet{pmlr-v28-wang13}. Then

\begin{equation}
\hat{r}_n^{(k)} \stackrel{p}{\to} r_n^{(k)}
\end{equation}

*Proof of Lemma 5.1*. 
This falls directly out of Theorem 5 of \citet{rubindelanchy2017statistical}. 
Since $\hat{V}_n^{(k)} \stackrel{p}{\to} V_n^{(k)}$ and $r(\cdot)$ is smooth, 
$\hat{r}_n^{(k)} \stackrel{p}{\to} r_n^{(k)}$.

**Lemma 5.2**. 
Let $\mu_n^{(k)} = \mu(V_n^{(k)})$ be the subspace incoherence of $V_n^{(k)}$ 
as defined by \citet{soltanolkotabi2012} (also note that 
$\mu(\hat{V}_n^{(k)}) = \mu(V_n^{(k)})$ using the extension of this definition 
by \citet{pmlr-v28-wang13}).

\begin{equation}
\mu_n^{(k)} = 0
\end{equation}

with probability 1 $\forall n, k$.

*Proof of Lemma 5.2*. 
Since $V_n^{(k)}$ is orthogonal to $V_n^{(l)}$ $\forall k \neq l$ (Lemma 3.1), 
this falls directly out of Theorem 2.8 of \citet{soltanolkotabi2012}.

**Lemma 5.3**. 
Let $(v_n^{(i)})^\top$ and $(\hat{v}_n^{(i)})^\top$ be the rows of $V_n$ and
$\hat{V}_n$ respectively. By \cite{rubindelanchy2017statistical},

\begin{equation}
\delta_n = \max_i ||\hat{v}_n^{(i)} - v_n^{(i)}|| \stackrel{p}{\to} 0
\end{equation}

**Proof of Theorem 5**. 
For ease of proving, we will consider $V_n$ and $\hat{V}_n$ instead of $Z_n$ 
and $\hat{Z}_n$.

By (12) and (13),

\begin{equation}
r(\hat{V}_n^{(k)}) - \mu(\hat{V}_n^{(k)}) = r_n^{(k)} \stackrel{p}{\to} r_n^{(k)} > 0
\end{equation}

Then by Theorem 6 of \citet{pmlr-v28-wang13}, (14) and (15) imply that  
$\Pr(\lim\limits_{n \to \infty} \text{ SSC detection property holds for } \hat{V}_n) = 1 - \epsilon$.
